Log-likelihood for summary data

### 2025-01-10 data-raw/pulling_oral_iv_data_cvtdb.Rmd

quick_stats <- function(.data) {
  stopifnot(is.data.frame(.data))
  cli::cli_h2("Statistics on the current data")
  message(paste0("Total number of data points: ", nrow(.data)))
  message("Unique Species and data counts:")
  print(table(.data$species))
  message(paste0("Number of unique studies: ", length(unique(.data$fk_study_id))))
  message(paste0("Number of unique series: ", length(unique(.data$fk_series_id))))
  message(paste0("Number of unique chemicals: ", length(unique(.data$analyte_dtxsid))))
  message(paste0("Unique Dose units: ", paste(unique(.data$dose_level_units_original), collapse = ", ")))
  message(paste0("Unique Concentration units: ", paste(unique(.data$conc_units_original), collapse = ", ")))
  message("Data by unique route of administration:")
  print(table(.data$administration_route_normalized))
  message("Data by medium collected:")
  print(table(.data$conc_medium_normalized))
  message("Data by curation set:")
  print(table(.data$curation_set_tag))
}

@@

# Harmonize values for easier analysis
cvt_df <- sql_query %>%
  mutate(
    across(ends_with("id"), as.character),
    # Note that sometimes fk_test_chemical_id is NA
    across(all_of(c("time_original",
                    "time_hr",
                    "conc_original",
                    "conc_sd_original",
                    "conc",
                    "conc_sd",
                    "dose_level_original",
                    "dose_level_normalized",
                    "loq")), as.numeric),
    # conc, conc_sd_original, and conc_original have NA/NE/ND values that are coerced to NAs
    year = as.character(year),
    species = tolower(species),
    sex = tolower(sex),
    analyte_name_original = tolower(analyte_name_original),
    dose_volume_units = stringr::str_extract(dose_volume,
                                    pattern = "(?<=[:space:]).+"),
    dose_volume = as.numeric(stringr::str_extract(dose_volume,
                                         pattern = "^(\\d|\\.)*")),
    dose_duration_units = stringr::str_extract(dose_duration,
                                      pattern = "(?<=[space]).+"),
    dose_duration = as.numeric(stringr::str_extract(dose_duration,
                                pattern = "^\\d")),
    radiolabeled = ifelse(radiolabeled == 1, TRUE, FALSE),
    n_subjects_normalized = ifelse(n_subjects_in_series %in% "NR", NA_real_,
      as.numeric(stringr::str_extract(n_subjects_in_series,
                                                 pattern = "\\d+")))
  )

@@

glimpse(cvt_df, width = 80)
```

@@

```{r}
quick_stats(cvt_df)
```

>>>>

quick_stats <- function(.data) {
  stopifnot(is.data.frame(.data))
  cli::cli_h2("Statistics on the current data")
  message(paste0("Total number of data points: ", nrow(.data)))
  message("Unique Species and data counts:")
  print(table(.data$species))

  message(paste0("Number of unique studies: ",
                 length(unique(.data$fk_study_id))))
  message(paste0("Number of unique series: ",
                 length(unique(.data$fk_series_id))))
  message(paste0("Number of unique chemicals: ",
                 length(unique(.data$analyte_dtxsid))))
  message(paste0("Unique Dose units: ",
                 toString(unique(.data$dose_level_units_original))))
  message(paste0("Unique Concentration units: ", 
                 toString(unique(.data$conc_units_original))))
  message("Data by unique route of administration:")
  print(table(.data$administration_route_normalized))
  message("Data by medium collected:")
  print(table(.data$conc_medium_normalized))
  message("Data by curation set:")
  print(table(.data$curation_set_tag))
}

@@

# Harmonize values for easier analysis
cvt_df <- sql_query %>%
  mutate(
    across(ends_with("id"), as.character),
    # Note that sometimes fk_test_chemical_id is NA
    across(all_of(c("time_original",
                    "time_hr",
                    "conc_original",
                    "conc_sd_original",
                    "conc",
                    "conc_sd",
                    "dose_level_original",
                    "dose_level_normalized",
                    "loq")), as.numeric),
    # conc, conc_sd_original, and conc_original
    # have NA/NE/ND values that are coerced to NAs
    year = as.character(year),
    species = tolower(species),
    sex = tolower(sex),
    analyte_name_original = tolower(analyte_name_original),
    dose_volume_units = stringr::str_extract(dose_volume,
                                    pattern = "(?<=[:space:]).+"),
    dose_volume = as.numeric(stringr::str_extract(dose_volume,
                                         pattern = "^(\\d|\\.)*")),
    dose_duration_units = stringr::str_extract(dose_duration,
                                      pattern = "(?<=[space]).+"),
    dose_duration = as.numeric(stringr::str_extract(dose_duration,
                                pattern = "^\\d")),
    radiolabeled = ifelse(radiolabeled == 1, TRUE, FALSE),
    n_subjects_normalized = ifelse(n_subjects_in_series %in% "NR", NA_real_,
      as.numeric(stringr::str_extract(n_subjects_in_series,
                                                 pattern = "\\d+")))
  )

@@

glimpse(cvt_df, width = 80)
quick_stats(cvt_df)
```

@@

```{r diff_log_function}
# Now I want to write a quick function that will log relevant changes
# from previous 'cvt' objects
devtools::load_all() # So this loads the previous cvt object
diff_log_cvt <- function(new_cvt,
                         key_id = "conc_time_id",
                         grp_ids = c("analyte_dtxsid",
                                     "species")) {
  stopifnot(
    key_id %in% colnames(cvt),
    key_id %in% colnames(new_cvt),
    all(grp_ids %in% colnames(cvt)),
    all(grp_ids %in% colnames(new_cvt))
  )
  if (length(unique(new_cvt[[key_id]])) != NROW(new_cvt)) {
    stop("key_id is not not a column that uniquely identifies this data.")
  }

  # Prepare some vectors:
  grp_strings_new <- unique(do.call("paste", new_cvt[grp_ids]))
  grp_strings_old <- unique(do.call("paste", cvt[grp_ids]))

  msg_to_write <- paste0("\n\n======== ", round(Sys.time()))

  if (identical(new_cvt, cvt)) {
    msg_to_write <- paste(msg_to_write, "No changes.", sep = "\t")
    message("There are no changes from previous `cvt` data.frame, ",
            "are you sure you want to save this?")
  } else {
    msg_to_write <- paste(msg_to_write, "Changes detected.", sep = "\t")

    ###---- Check key_id changes (usually conc_time_id)
    diff_row_add <- setdiff(new_cvt[[key_id]], cvt[[key_id]])
    diff_row_sub <- setdiff(cvt[[key_id]], new_cvt[[key_id]])
    diff_row_common <- intersect(cvt[[key_id]], new_cvt[[key_id]])
    # Nature of data changes:
    added_grps <- unique(
      do.call("paste",
              new_cvt[which(new_cvt[[key_id]] %in% diff_row_add), grp_ids]
              )
      )
    removed_grps <- unique(
      do.call("paste",
              cvt[which(cvt[[key_id]] %in% diff_row_sub), grp_ids]
              )
      )

    added_new_grps <- setdiff(added_grps, grp_strings_old)
    added_existing_grps <- intersect(added_grps, grp_strings_old)
    removed_new_grps <- setdiff(removed_grps, grp_strings_new)
    removed_existing_grps <- intersect(removed_grps, grp_strings_new)

    msg_to_write <- c(
      msg_to_write,
      paste("Rows added:", length(diff_row_add)),
      paste("Rows removed:", length(diff_row_sub))
    )

    limit_grp_output <- function(grp_vector) {
      if (length(grp_vector) > 0) {
        if (length(grp_vector) > 6) {
          grp_vector <- c(grp_vector[c(1,2)],
                              paste0("... [",
                                     length(grp_vector) - 3,
                                     "] more!"),
                              grp_vector[c(length(grp_vector) - 1)]
          )
        }
      return(grp_vector)
    }
      msg_to_write <- c(msg_to_write,
                        "New Groups added (not in previous data):",
                        paste0("\t", limit_grp_output(added_new_grps)),
                        "Groups with new observations (group present in previous data)",
                        paste0("\t", limit_grp_output(added_existing_grps)),
                        "Groups removed that no longer have data present:",
                        paste0("\t", limit_grp_output(removed_new_grps)),
                        "Groups removed that still have some data present:",
                        paste0("\t", limit_grp_output(removed_existing_grps)),
                        "-----------------------------------"
                        )
    }

    ### -----Checking columns
    diff_col_add <- setdiff(names(new_cvt), names(cvt)) # Columns added in new_cvt
    diff_col_sub <- setdiff(names(cvt), names(new_cvt)) # Columns removed in new_cvt
    diff_col_common <- intersect(names(cvt), names(new_cvt))

    # Left off here TODO
    if (length(c(diff_col_add, diff_col_sub)) == 0) {
      msg_to_write <- c(msg_to_write,
                        "No changes in column names.")
    } else if (length(diff_col_add) > 0 && length(diff_col_sub) > 0) {
      # Check renaming
      common_row_cvt <- cvt[diff_row_common, ]
      common_row_new <- new_cvt[diff_row_common, ]

      diff_columns <- function(x, y) {
        out_list <- character()
        for (ix in names(x)) {
          for (iy in names(y)) {
            if (identical(x[[c(key_id,ix)]], y[[c(key_id, iy)]])) {
              out_list <- c(out_list, paste0(ix, " -> ", iy))
            }
          }
        }
        return(out_list)
      }

      renamed_cols <- diff_columns(common_row_cvt[diff_col_sub],
                                   common_row_new[diff_col_add])
      if (length(renamed_cols) > 0) {
        msg_to_write <- c(msg_to_write,
                          "Renamed columns (non-exhaustive):",
                          renamed_cols)
      } else {
        msg_to_write <- c(msg_to_write,
                          "No renamed columns found.")
      }
    }
    if (length(diff_col_add) > 0) {
      msg_to_write <- c(msg_to_write,
                        "--------------",
                        "Columns added:",
                        diff_col_add)
    }
    if (length(diff_col_sub) > 0) {
      msg_to_write <- c(msg_to_write,
                        "______________",
                        "Columns removed (from previous data):",
                        diff_col_sub)
    }

    if (!identical(common_row_cvt[diff_col_common],
                   common_row_new[diff_col_common])) {
      msg_to_write <- c(msg_to_write,
                        "Data for same observation key_ids have changed.")
    }
    msg_to_write <- c(msg_to_write,
                      "================================\n")
  }
  return(cat(msg_to_write, sep = "\n"))
}
```

### 2025-01-10 data-raw/summary_loglike_lognormal.Rmd

```{r}
get_sample_moments <- function(meanlog, sdlog, n_samp){
  #samples
  y <- rlnorm(n = n_samp, meanlog = meanlog, sdlog = sdlog)
  #log-transformed samples
  ylog <- log(y)
  return(c("sample_mean" = mean(y),
           "sample_sd" = sd(y),
           "sample_meanlog" = mean(ylog),
           "sample_sdlog" = sd(ylog),
           "sample_n" = n_samp))
}
```

@@

```{r}
samp[c("est_sample_meanlog",
       "est_sample_sdlog")] <- convert_summary_to_log(sample_mean = samp$sample_mean,
                                                                   sample_SD = samp$sample_sd)
```

@@

Log-scale sample SD. The "true" population log-scale SD (1.0) is indicated by vertical and horizontal dashed lines.

```{r}
ggplot(data = samp) +
  geom_point(aes(x = sample_sdlog,
                 y = est_sample_sdlog),
             color = "gray50") +
  geom_abline(intercept = 0, slope = 1, color = "black") +
    geom_hline(yintercept = 1, linetype = 2) +
  geom_vline(xintercept = 1, linetype = 2) +
  xlab("Actual log-scale sample SD") +
  ylab("Estimated log-scale sample SD") +
  ggtitle("Estimated vs. actual log-scale sample SD, n = 7") +
  theme_bw()
```

The estimated log-scale sample SD usually underestimates the actual log-scale sample SD (most points are below the identity line).

>>>>

```{r}
get_sample_moments <- function(meanlog, sdlog, n_samp){
  #samples
  y <- rlnorm(n = n_samp, meanlog = meanlog, sdlog = sdlog)
  #log-transformed samples
  ylog <- log10(y)
  return(c("sample_mean" = mean(y),
           "sample_sd" = sd(y),
           "sample_meanlog" = mean(ylog),
           "sample_sdlog" = sd(ylog),
           "sample_n" = n_samp))
}
```

@@

```{r}
samp[c("est_sample_meanlog",
       "est_sample_sdlog")] <- convert_summary_to_log10(sample_mean = samp$sample_mean,
                                                                   sample_SD = samp$sample_sd)
```

@@

Sampling error:

```{r}
ggplot(data = samp) +
  geom_histogram(aes(x = est_sample_meanlog - sample_meanlog))
```

Log-scale sample SD. The "true" population log-scale SD (`log10(exp(1))`) is indicated by vertical and horizontal dashed lines.

```{r}
ggplot(data = samp) +
  geom_point(aes(x = sample_sdlog,
                 y = est_sample_sdlog),
             color = "gray50") +
  geom_abline(intercept = 0, slope = 1, color = "black") +
    geom_hline(yintercept = log10(exp(1)), linetype = 2) +
  geom_vline(xintercept = log10(exp(1)), linetype = 2) +
  xlab("Actual log-scale sample SD") +
  ylab("Estimated log-scale sample SD") +
  ggtitle("Estimated vs. actual log-scale sample SD, n = 7") +
  theme_bw()
```

The estimated log-scale sample SD usually overestimates the actual log-scale sample SD (most points are above the identity line).
